---
title: "Smith Creek near New Market, VA"
author: "Dave Holtschlag"
date: "January 18, 2018"
output: html_document
---

The analysis involves forecast NO2- + NO3-N concentrations at Smith Creek near New Market, Va. 


```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(GGally)
# scales implements function boxcox_trans
library(scales)
library(MASS)
library(mgcv)
library(astsa)
knitr::opts_chunk$set(echo = TRUE)
```

### Read streamflow and water quality data for Smith Creek

Daily flow and water-quality data for 01632900 Smith Creek near New Market, VA, was retrieved from NWIS. The resulting data file was stored as 01632900.txt.  This code chunk reads and renames selected fields labelled with parameter code and statistic codes to more familiar codes.  None included fields include agency_cd, site_no, and field associated with minimum or maximum daily values. The daily median nitrate nitrogen concentration (NO23_MEDI) was considered the target variable. 

```{r read_data}
df <- read_tsv(file = '../data/01632900.txt', comment = '#')

sta_no   <- '01632900'
sta_name <- 'Smith Creek near New Market, VA'

df1 <- df %>% 
   rename(flow_mean = `144686_00060_00003`,
          flow_code = `144686_00060_00003_cd`,
          temp_medi = `144689_00010_00008`,
          temp_code = `144689_00010_00008_cd`,
          sc25_medi = `144692_00095_00008`,
          sc25_code = `144692_00095_00008_cd`,
          pH_medi   = `144695_00400_00008`,
          pH_code   = `144695_00400_00008_cd`,
          DO_medi   = `144701_00300_00008`,
          DO_code   = `144701_00300_00008_cd`,
          NO23_medi = `144704_99133_00008`,
          NO23_code = `144704_99133_00008_cd`,
          turb_medi = `214378_63680_00008`,
          turb_code = `214378_63680_00008_cd`) %>% 
   dplyr::select(datetime,
                 flow_mean, flow_code, temp_medi, temp_code,
                 sc25_medi, sc25_code, pH_medi,   pH_code,  
                 pH_medi,   pH_code,   DO_medi,   DO_code,
                 NO23_medi, NO23_code, turb_medi, turb_code)

```

### Time series plots of selected parameters


```{r time_plots}

# Streamflow
df1 %>% 
   ggplot( aes(x = datetime, y = flow_mean, color = flow_code)) +
   geom_point() +
   theme_few()  +
   geom_vline( xintercept = df1$datetime[which(is.na(df1$flow_mean))], 
               color = 'grey', alpha = 0.1) +
   scale_y_continuous( trans = 'log10', limits = c(5, 5000), 
                       breaks = c(5, 10, 50, 100, 500, 1000, 5000)) +
   labs(x = 'YEAR', y = 'FLOW, IN CUBIC FEET PER SECOND',
        title = paste('Daily Mean Streamflow at', sta_no, sta_name)) +
   theme( legend.position = 'bottom')


df1 %>% 
   ggplot( aes(x = datetime, y = temp_medi, color = temp_code)) +
   geom_point() +
   theme_few()  + 
   geom_vline( xintercept = df1$datetime[which(is.na(df1$temp_medi))], 
               color = 'grey', alpha = 0.5) +
   scale_y_continuous( trans = 'identity', limits = c(0, 30)) +
   labs(x = 'YEAR', y = 'TEMPERATURE, IN DEGREES CELCIUS',
        title = paste('Daily Median Water Temperature at', sta_no, sta_name)) +
   theme( legend.position = 'bottom')


# Find range of lambda parameters to normalize specific conductance at 25 degC
boxcox_sc25 <- boxcox(df1$sc25_medi ~ 1, lambda = seq(4,5, 1/10))
# Find lambda with maximum likelihood
ndx_max <- which.max(boxcox_sc25$y)

# boxcox (bc) transformation
df1$sc25_bcstd <- ( df1$sc25_medi^boxcox_sc25$x[ndx_max] - 1 ) / boxcox_sc25$x[ndx_max]
# standardizing (std) transformation 
df1$sc25_bcstd <- (df1$sc25_bcstd - mean(df1$sc25_bcstd, na.rm = TRUE ) ) / 
   sd(df1$sc25_bcstd, na.rm = TRUE)

# Transformed Specific Conductance 
df1 %>% 
   ggplot( aes(x = datetime, y = sc25_medi, color = sc25_code)) +
   geom_point() +
   theme_few()  +
   geom_vline( xintercept = df1$datetime[which(is.na(df1$sc25_medi))], 
               color = 'grey', alpha = 0.5) +
   geom_smooth( method = 'lm') +
   scale_y_continuous( trans = boxcox_trans(p=boxcox_sc25$x[ndx_max]) ) +
   labs(x = 'Year', y = expression(paste('Specific conductance,in  ',mu,'-siemens/cm at '*25~degree*C)), 
        title = paste('Specific conductance at ', sta_no, sta_name)) +
   theme( legend.position = 'bottom') 

rlm_sc25_bcstd <- rlm(df1$sc25_bcstd[1:1995] ~ df1$datetime[1:1995])
print(summary(rlm_sc25_bcstd))


# Plot of pH
df1 %>% 
   ggplot( aes(x = datetime, y = pH_medi, color = pH_code)) +
   geom_jitter( height = 0.03) +
   theme_few()  +
   geom_vline( xintercept = df1$datetime[which(is.na(df1$pH_medi))], 
               color = 'grey', alpha = 0.5) +
   scale_y_continuous( trans = 'identity') +
   labs(x = 'YEAR', y = 'pH', 
        subtitle = 'Note: Jitter added to pH, which is rounded to 0.1 units.',
        title = paste('pH at ', sta_no, sta_name)) +
   theme( legend.position = 'bottom') 

# Plot Dissolved Oxygen
df1 %>% 
   ggplot( aes(x = datetime, y = DO_medi, color = DO_code)) +
   geom_point() +
   theme_few()  +
   geom_vline( xintercept = df1$datetime[which(is.na(df1$DO_medi))], 
               color = 'grey', alpha = 0.5) +
   scale_y_continuous( trans = 'identity') +
   labs(x = 'YEAR', y = 'Dissolved oxygen, in mg/L', 
        title = paste('Daily Median Dissolved Oxygen at ', sta_no, sta_name)) +
   theme( legend.position = 'bottom') 

# Plot of Nitrate plus nitrite nitrogen
df1 %>% 
   ggplot( aes(x = datetime, y = NO23_medi, color = NO23_code)) +
   geom_point() +
   geom_smooth( method = 'rlm' ) +
   theme_few()  +
   geom_vline( xintercept = df1$datetime[which(is.na(df1$NO23_medi))], 
               color = 'grey', alpha = 0.5) +
   scale_y_continuous( trans = 'identity') +
   labs(x = 'Year', y = expression(~NO[3]~ + ~NO[2]~ 'Nitrogen, in mg/L' ), 
        title = paste(expression('Daily Median' ~NO[3]~ + ~NO[2]~ 'Nitrogen at'),sta_name) ) +
   theme( legend.position = 'bottom') 


# Note: Data for rows 1:1995 includes only the archived data (not provisional)
lm1 <- lm(NO23_medi ~ datetime, data = df1[1:1995,])

print(summary(lm1))

# Note: Data for rows 1:1995 includes only the archived data (not provisional)
rlm1 <- rlm(NO23_medi ~ datetime, data = df1[1:1995,])

print(summary(rlm1))

# Plot of turbidity
df1 %>% 
   ggplot( aes(x = datetime, y = turb_medi, color = turb_code)) +
   geom_point() +
   theme_few()  + 
   geom_vline( xintercept = df1$datetime[which(is.na(df1$turb_medi))], 
               color = 'grey', alpha = 0.5) +
   scale_y_continuous( trans = 'log10', limits = c(0.1, 500),
                       breaks = c(0.1, 0.5, 1, 5, 10, 50, 100, 500) ) +
   labs(x = 'Year', y = paste('Turbidity, FNU'), 
        title = paste('Daily Median Turbidity at',sta_no, sta_name ) ) +
   theme( legend.position = 'bottom') 

```


### Multivariate associations among parameters

```{r plot_matrix, warning = FALSE, fig.width = 9, fig.height = 8}

df1 %>% 
   mutate(log_flow = log10(flow_mean),
          log_turb = log10(turb_medi)) %>% 
   ggpairs(columns = c('log_flow', 'temp_medi', 'sc25_medi', 'pH_medi', 'DO_medi',
                       'log_turb', 'NO23_medi'),
           lower = list(continuous = 'points', alpha = 0.2, size = 0.02),
           title = 'Relations among parameters at Smith Creek near New Market, Va.') +
   theme_few() 


```


```{r load_data, eval = TRUE}
# Define day-of-year for seasonal function
df1$doy      <- as.numeric(format(df1$datetime, '%j'))
df1$delT     <- as.numeric(df1$datetime - df1$datetime[1] + 1)
df1$log_flow <- log10(df1$flow_mean)
df1$dlq      <- c(NA,diff(df1$log_flow))
df1$sin1_doy <- sin(2 * pi * df1$doy / 366)
df1$sin2_doy <- sin(2*2 * pi * df1$doy / 366)
df1$cos1_doy <- cos(2 * pi * df1$doy / 366)
df1$cos2_doy <- cos(2*2 * pi * df1$doy / 366)

load(file = "C:/Home/Projects/NawqaIWS/QW_forecast/sea01632900.RData")
df1 <- left_join(df1, sea01632900, by = 'doy')


# Use subset that excludes provisional NO23 data
df2          <- df1[1:1995, ]
```


```{r model_ts}
# Model series as AR(1) process
ar1_logQ  <- sarima(df2$log_flow, 1, 0, 0) 
print(ar1_logQ)

# Model log_flow as an arma(2,0,1) process
ar2_logQ  <- sarima(df2$log_flow, 2, 0, 1) 
print(ar2_logQ)

# Model log_flow as an ar_1_lmsea component where lmsea was based on a lm using period of record
ar1_logQ_lmsea <- sarima(df2$log_flow, 1, 0, 0, 
                      xreg = cbind(df2$sincos))
print(ar1_logQ_lmsea)

# Model log_flow as an ar_1gamsea component where gamsea was based on a gam using period of record
ar1_logQ_gamsea <- sarima(df2$log_flow, 1, 0, 0, 
                      xreg = cbind(df2$gamsea))
print(ar1_logQ_gamsea)

# Model log_flow as an ar_1_sincos where sincos components are based on sample period
ar1_logQ_sincos <- sarima(df2$log_flow, 1, 0, 0,
                          xreg = cbind(df2$sin1_doy, df2$cos1_doy))

print(ar1_logQ_sincos)

# Conclude: sin and cos components for sampled doy looks adequate and practical

df2$log_flow_pred <- df2$log_flow - ar1_logQ_sincos$fit$residuals

df2 %>% 
   ggplot( aes( x = datetime)) +
   geom_line(aes(y = log_flow), color = 'blue') +
   geom_line(aes(y = log_flow_pred), color = 'red') +
   # scale_x_date(limits = as.Date(c('01/01/2014', '01/01/2018'), format = '%d/%m/%Y')) +
   labs( x = 'Year', y = 'log10 Daily Flow, in ft^3/s',
         title = 'Relation between measured and predicated flow at 01632900 Smith Creek near New Market, Va.') +
   geom_vline( xintercept = df2$datetime[which(is.na(df2$flow_mean))], 
               color = 'grey', alpha = 1) +
   theme_few()

```



```{r sarima_NO23, fig.width = 9}

# AR(1) without seasonal component
ar100_NO23     <- sarima(df2$NO23_medi, 1, 0, 0 ) 
print(ar100_NO23) 
# AICc = -3.198275

# AR(1) with sampled seasonal component
ar100_NO23_sincos <- sarima(df2$NO23_medi, 1, 0, 0, 
                      xreg = cbind(df2$sin1_doy, df2$cos1_doy) )
print(ar100_NO23_sincos)
# AICc: -3.19685

ar100_NO23_sc  <- sarima(df2$NO23_medi, 1, 0, 0, 
                      xreg = df2$sc25_bcstd ); 
print(ar100_NO23_sc )
# AICc: -3.573532 -- Better than model w or w/o sampled seasonal component
df2$ar100_NO23_sc_pred <- df2$NO23_medi - ar100_NO23_sc$fit$residuals
acf2(ar100_NO23_sc$fit$residuals)

ar101_NO23_sc  <- sarima(df2$NO23_medi, 1, 0, 1, 
                      xreg = df2$sc25_bcstd ); 
print(ar101_NO23_sc )
acf2(ar101_NO23_sc$fit$residuals)
df2$ar101_NO23_sc_pred <- df2$NO23_medi - ar101_NO23_sc$fit$residuals

# NO23 concentration with ar(2) and x = sc25bcstd
ar200_NO23_sc  <- sarima(df2$NO23_medi, 2, 0, 0, 
                      xreg = df2$sc25_bcstd ); 
print(ar200_NO23_sc )
acf2(ar200_NO23_sc$fit$residuals)
df2$ar200_NO23_sc_pred  <- df2$NO23_medi - ar200_NO23_sc$fit$residuals
df2$ar200_NO23_sc_resid <- ar200_NO23_sc$fit$residuals

df2 %>% 
   ggplot( aes( x = datetime)) +
   geom_vline( xintercept = df2$datetime[which(is.na(df2$NO23_medi))], 
               color = 'grey', alpha = 1) +
   geom_line(aes(y = NO23_medi), color = 'blue') +
   geom_line(aes(y = ar200_NO23_sc_pred), color = 'red') +
   # scale_x_date(limits = as.Date(c('01/01/2014', '01/01/2018'), format = '%d/%m/%Y')) +
   labs( x = 'Year', y = 'log10 Daily Flow, in ft^3/s',
         title = 'Time series of measured and predicated flow at 01632900 Smith Creek near New Market, Va.') +
   theme_few()


df2 %>% 
   ggplot( aes( x = datetime, y = ar200_NO23_sc$fit$residuals)) +
   geom_vline( xintercept = df2$datetime[which(is.na(df2$NO23_medi))], 
               color = 'grey', alpha = 1) +
   geom_point() +
   theme_few() +
   geom_hline( yintercept = 0, color = 'red', linetype = 'dashed')

df2 %>% 
   ggplot( aes(x = ar200_NO23_sc$fit$residuals)) +
   geom_density() +
   stat_function(fun = dnorm, args = list(mean = mean(ar200_NO23_sc$fit$residuals, na.rm = TRUE), 
                                          sd = sd(ar200_NO23_sc$fit$residuals, na.rm = TRUE)), 
                 color = 'red', linetype = 'solid') +
   theme_few() +
   theme()     +
   labs( x = 'NO3- + NO2-N Residuals from Model', y = 'Probability Density',
         title = paste('Distribution of NO3- + NO2-N residuals from AR(2,0)X(sc25_bcstd) at Streamgage ', sta_no) )


y1 <- quantile(ar200_NO23_sc$fit$residuals[!is.na(ar200_NO23_sc$fit$residuals)], c(0.25, 0.75))
x1 <- qnorm( c(0.25, 0.75) )
slope <- diff(y1)/diff(x1)
int   <- y1[1L] - slope * x1[1L]

df2 %>%
   ggplot( aes( sample = ar200_NO23_sc$fit$residuals)) +
   geom_qq(geom = "point", distribution = stats::qnorm,
           na.rm = FALSE) +
   theme_few() +
   geom_abline( intercept = int, slope = slope, color = 'blue', linetype = 'dashed') +
   labs( x = 'Standard Normal Quantiles', y = 'AR(2) NO3- + NO2-N Residual Quantiles',
         title = 'Heavy Tailed Distribution of AR(2)X(sc25) Residuals Relative to Normal')


save(df2, file = 'df2_AR2Xsc25.RData')

```



```{r next_step, eval = FALSE}


sar_200     <- sarima(df2$NO23_medi, 2, 0, 0); print(sar_200) # AICc = -3.229706  * 

sar_201     <- sarima(df2$NO23_medi, 2, 0, 1); print(sar_201) # AICc = -3.22893

# Trend
sar_200_trnd   <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$delT))
print(sar_200_trnd) # AICc = -3.22999 *, df = 1828

# Flow
sar_200_lflow  <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$log_flow))
print(sar_200_lflow) # AICc = -3.229706, df = 1828

# Water temperature
sar_200_temp   <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$temp_medi))  
print(sar_200_temp)  # AICc = -3.232070, df = 1813

# sc25 indicates the bc and std transformed 
sar_200_sc25  <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$sc25_bcstd))
print(sar_200_sc25) # AICc = -3.636874, df = 1772 

# pH_medi
sar_200_pHmedi <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$pH_medi))
print(sar_200_pHmedi) # AICc = -3.335286, df = 1808

# DO_medi
sar_200_DOmedi <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$DO_medi))
print(sar_200_DOmedi) # AICc = -3.245214, df = 1806

# log10(turb_medi)
sar_200_lnTurb <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$log_turb))
print(sar_200_lnTurb) # AICc = -3.229706, df = 1829

# flow and sc25
sar_200_FlowSc25 <- sarima(df1$NO23_medi, 2, 0, 0, xreg = cbind(df1$log_flow, df1$sc25_bcstd)) 
print(sar_200_FlowSc25) # AICc = -3.577086, df = 1899



# AICc = -3.206867 sin + cos term adds nothing
sar_200_sin_cos <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$sin_doy, df2$cos_doy)) 
print(sar_200_sin_cos) # AICc = -3.228434

# flow and trnd
sar_200_FlowTrnd <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$log_flow, df2$delT)) 
print(sar_200_FlowTrnd) # AICc = -3.22999, df = 1828

# sc25 and trnd
sar_200_TrndSc25 <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$sc25_bcstd, df2$delT)) 
print(sar_200_TrndSc25) # AICc = -3.636543, df = 1899

# Preferred Model is sar_200_sc25 AKA ar2SC_NO23
# sc25 indicates the bc and std transformed 
ar2SC_NO23           <- sarima(df2$NO23_medi, 2, 0, 0, xreg = cbind(df2$sc25_bcstd))
print(ar2SC_NO23) # AICc = -3.636874, df = 1772 
df2$ar2SC_NO23_resid <- ar2SC_NO23$fit$residuals
df2$ar2SC_NO23_pred  <- df2$NO23_medi - df2$ar2SC_NO23_resid


```

### Plot results for AR(2) Model


```{r plot_results, eval = FALSE}

# Relation between measured and predicted NO3- + NO2-N concentrations at Smith Cr., VA 
df2 %>% 
   ggplot( aes(x = NO23_medi, y = ar2SC_NO23_pred)) +
   geom_point() + 
   theme_few()  +
   geom_abline( intercept = 0, slope = 1, color = 'red', linetype = 'dashed') +
   labs(x = 'Measured Nitrate + Nitrite Nitrogen, in mg/L', 
        y = 'AR(2) Predicted Nitrate + Nitrite Nitrogen',
        title = 'Relation between measured and predicted NO3 + NO2 at Smith Creek, Va.')

# Time series of residual NO3- + NO2-N concentrations at Smith Cr., VA 
df2 %>% 
   ggplot( aes(x = datetime, y = ar2SC_NO23_resid)) +
   geom_point() + 
   theme_few()  +
   geom_hline( yintercept = 0, color = 'red', linetype = 'dashed') +
   labs(x = 'Year', 
        y = 'Residual AR(2) Nitrate + Nitrite Nitrogen, in mg/L',
        title = 'Daily Time Series of NO3- + NO2-N Concentrations at Smith Creek, Va.')

```


```{r resid_den, eval = FALSE}

df2 %>% 
   ggplot( aes(x = ar2SC_NO23_resid)) +
   geom_density() +
   stat_function(fun = dnorm, args = list(mean = mean(df2$ar2SC_NO23_resid, na.rm = TRUE), 
                                          sd = sd(df2$ar2SC_NO23_resid, na.rm = TRUE)), 
                 color = 'red', linetype = 'solid') +
   theme_few() +
   theme()     +
   labs( x = 'NO3- + NO2-N Residuals from AR(2) model', y = 'Probability Density',
         title = paste('Distribution of NO3- + NO2-N residuals from AR(2) at Streamgage ', sta_no) )


y1 <- quantile(df2$ar2SC_NO23_resid[!is.na(df2$ar2SC_NO23_resid)], c(0.25, 0.75))
x1 <- qnorm( c(0.25, 0.75) )
slope <- diff(y1)/diff(x1)
int   <- y1[1L] - slope * x1[1L]

df2 %>% 
   ggplot( aes( sample = ar2SC_NO23_resid)) +
   geom_qq(geom = "point", distribution = stats::qnorm,
           na.rm = FALSE) +
   theme_few() + 
   geom_abline( intercept = int, slope = slope, color = 'red', linetype = 'dashed') +
   labs( x = 'Standard Normal Quantiles', y = 'AR(2) NO3- + NO2-N Residual Quantiles',
         title = 'Heavy Tailed Distribution of AR(2) Residuals Relative to Normal')


```


```{r model_acf, eval = FALSE}

# Autocorrelation Function
acf(df2$ar2SC_NO23_resid, na.action = na.pass)
title('\nACF of AR(2) NO3- + NO2-N Residuals at Smith Creek')

# Partial autocorrelation Function
acf(df2$ar2SC_NO23_resid, na.action = na.pass, type = 'partial')
title('\nPACF of AR(2) NO3- + NO2-N Residuals at Smith Creek')

```

```{r save_model, eval = FALSE}

# Save df2 and selected model object
save(df2, ar2SC_NO23, file = 'ar2SC_NO23.RData')


```
